---
title: "heatflux.Rmd"
author: "Leeya Pressburger"
date: "2/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Processing ocean heat flux data from CMIP6

First, I read in the CMIP6 csv files that I accessed with Pangeo in Python and 
created an output data frame for all six heat flux variables.

``` {r intro}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

BASE_DIR <- here::here()

# Variables of interest
vars <- c("hfls", "hfss", "rlds", "rlus", "rsds", "rsus")

# Function to get lists of file paths for each of the six variables
get_path <- function(var){
  assign(x = paste0("path_", var), 
         value = file.path(BASE_DIR, "heat_flux", var))
}

# Access file paths for specified variables
paths <- lapply(vars, get_path)
paths <- c(paths[[1]], paths[[2]], paths[[3]], paths[[4]], paths[[5]], paths[[6]])

# Function to list the files for each path specified
list_files <- function(path){
  list.files(path = path, pattern = "*.csv", full.names = TRUE)
}

# List files for specified paths
files <- lapply(paths, list_files)

# For each file in each file path for each variable, read in the csv data
# Force column types
output <- list()
for(num in 1:length(files)){
  output[[num]] <- lapply(files[[num]], read_csv, col_types = "dccccccdd") %>%
    bind_rows(.id = "File")
}

# Create data frame
output <- bind_rows(output)

head(output)

```

Next, I added name and row number columns to more easily manipulate the data later.
I also corrected the years of the CO~2~ forcing experiments to have a start year of 1850. 

``` {r cleanup_data}
# Add name and row number identifiers to output dataframe
output$name <- paste0(output$model, "_", 
                      output$ensemble, "_", 
                      output$experiment)

output$rownum <- seq_len(nrow(output))

# Correct non-conventional years in CO2 forcing experiments
# Scale to start at 1850
output1 <- output %>%
  select("rownum", "year", "File") %>%
  filter(year < 1850 | year > 2500) %>%
  group_by(File) %>%
  mutate(start_year = year[1],
         scale = year - start_year,
         new_year = 1850 + scale) %>%
  select("rownum", "new_year")

# Combine with original dataframe
new_output <- left_join(output, output1, by = "rownum")

head(new_output)

```

This next part was a bit tricky. Because of Pangeo complications, there weren't the same number of files for each variable (ranging from 919 files to 926 files).
This chunk tries to isolate only the file names that all six variables share (pardon the hardcoding, I didn't feel like messing with a function/loop).

``` {r names}
# Need to get common names across six variables
names <- unique(new_output$name)

# How many model/experiment/ensemble combinations are there for each variable?
group_by_var <- new_output %>%
  group_by(variable) %>%
  summarize(names = unique(name)) %>%
  mutate(count = length(names))

counts <- group_by_var %>% summarize(variable, count) %>% unique()

# Which variables have the most and least names?
most_names <- group_by_var %>% filter(variable == "rsds")
hfls_names <- group_by_var %>% filter(variable == "hfls")
hfss_names <- group_by_var %>% filter(variable == "hfss")
rlds_names <- group_by_var %>% filter(variable == "rlds")
rlus_names <- group_by_var %>% filter(variable == "rlus")
rsus_names <- group_by_var %>% filter(variable == "rsus")

# Extract common names and names that are missing
common_names <- left_join(most_names, hfls_names, by = "names")
common_names <- left_join(common_names, hfss_names, by = "names")
common_names <- left_join(common_names, rlds_names, by = "names")
common_names <- left_join(common_names, rlus_names, by = "names")
common_names <- left_join(common_names, rsus_names, by = "names")
common_names <- drop_na(common_names)

good_names <- common_names$names

# Manually remove a name that is problematic later
good_names <- good_names[-337]

```

This chunk gets ready to manipulate the data to do the heat flux calculation.
It reorganizes the output data frame and selects only the necessary columns.

``` {r sort_columns}
# Reorganize data frame
data <- new_output %>%
  # Remove pesky NA variable
  filter(variable != "NA") %>%
  # Remove missing names
  filter(name %in% good_names) %>%
  # Organize
  select(-c("X1", "x", "File.x", "File.y")) %>%
  relocate(c("rownum", "name"), .before = "variable")

# Problem - need to make one year column with "year" and "new_year"
# If new_year has an NA, replace it with year, otherwise leave as is
replace_year <- ifelse(is.na(data$new_year), 
                data$year, 
                data$new_year)
data$year <- replace_year
# Get rid of new_year column
data <- data %>% select(-new_year)

head(data)
```

Now for the actual calculation.

``` {r equation}
# Heat flux equation
# rsds - rsus + rlds - rlus - hfss - hfls

# Select important columns
heat_flux <- data %>% select(c(name, model, experiment, year, variable, value))

# Reshape data
heat_flux <- heat_flux %>%
  group_by(name, year) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  ungroup()

# Compute equation
hf_output <- list()
for(n in seq_len(nrow(heat_flux))){
  hf_output[[n]] <- heat_flux$rsds[[n]] - heat_flux$rsus[[n]] + 
    heat_flux$rlds[[n]] - heat_flux$rlus[[n]] - 
    heat_flux$hfss[[n]] - heat_flux$hfls[[n]]
}

# Make output numeric, not a list
hf_output <- as.numeric(as.character(hf_output))

# Combine output with data frame
heat_flux <- heat_flux %>%
  mutate(value = hf_output)

head(heat_flux)

```

Now, comparing historical data from CMIP5 to CMIP6.

``` {r compare}
# Read in csv
hectorcal <- read.csv(file.path(BASE_DIR, "./heat_flux/CMIP5_heat_flux_final.csv"))
hc <- hectorcal %>% filter(experiment == "historical", model == "CanESM2")

hist <- heat_flux %>%
  filter(experiment == "esm-hist", model == "CanESM5")

summary(hist$value)
summary(hc$value)
```