---
title: "Processing tas over land"
author: "Leeya Pressburger"
date: "2/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Processing surface temperature over land

I followed a similar process here as I did with regular `tas`. First, I simply read in the appropriate csv files. 

``` {r start, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

BASE_DIR <- here::here()
path = paste0(BASE_DIR, "/tas_land")

files <- list.files(path = path, pattern = "*.csv", full.names = TRUE)

# Read in data, forcing column types
data <- lapply(files, read_csv, col_types = "dccccccdd") %>%
  bind_rows(.id = "File")

# Add name and row number identifier
data$name <- paste0(data$model, "_", data$ensemble, "_", data$experiment)
data$rownum <- seq_len(nrow(data))

head(data)

```

Next, I cleaned up the data set by shifting the CO2 forcing experiments to be relative to the year 1850. I then created a new dataset where each experiment corresponded to the corrected years.

``` {r years, message = FALSE}
# Correct non-conventional years in CO2 forcing experiments
# Scale to start at 1850
years <- data %>%
  select("rownum", "year", "File") %>%
  filter(year < 1850 | year > 2500) %>%
  group_by(File) %>%
  mutate(start_year = year[1],
         scale = year - start_year,
         new_year = 1850 + scale) %>%
  select("rownum", "new_year")

# Combine with original dataframe
tas <- left_join(data, years, by = "rownum")

# Problem - need to make one year column with "year" and "new_year"
# If new_year has an NA, replace it with year, otherwise leave as is
replace_year <- ifelse(is.na(tas$new_year), 
                       tas$year, 
                       tas$new_year)

tas$year <- replace_year

# Get rid of unnecessary columns
tas <- tas %>% select(c(-new_year, -File.y, -X1))

head(tas)

```

Next, I calculated `Tgav` by finding the models that had historical runs, calculating their historical averages, and then finding `Tgav` by subtracting the average from `tas`.

``` {r Tgav, message = FALSE}
# Get historical average
hist <- tas %>%
  filter(experiment == "esm-hist") %>%
  group_by(model) %>%
  mutate(hist_av = mean(value))

# Isolate models and their historical averages
historical <- as.data.frame(unique(hist$model))
historical$avg <- unique(hist$hist_av)
colnames(historical) <- c("model", "avg")

# Calculate Tgav
models <- historical$model
Tgav_land <- tas %>%
  filter(model %in% models, !experiment == "esm-hist") %>%
  left_join(historical, by = "model") %>%
  mutate(Tgav = value - avg)

head(Tgav_land)

```

Below are several graphs with the `land_tas` data. I first compared the models that have `Tgav` values to the `tas` data from global runs. I then plotted a few more specific graphs, including all land models against global, and all land models on their own.

There were a few models that returned unusually low values for `tas`. They are listed below.

``` {r graphs}

# Graph tas_global vs tas_land - only for models with Tgav values
# Assign type identifier
Tgav_land$type <- "land"

# Access only needed columns
Tgav_l_plot <- Tgav_land %>%
  select(c(year, value, model, experiment, ensemble, type))

# Using global data from "processing_tas.R"
Tgav <- read.csv("./tas/Global_tas_data.csv")
Tgav_plot <- Tgav %>%
  select(year, value, model, experiment, ensemble, type)

# Combine global and land values
plot_data <- bind_rows(Tgav_l_plot, Tgav_plot)

# Plot only models that have Tgav values
plot_data %>%
  # Removing an abnormally low model
  filter(model != "MPI-ESM1-2-LR") %>%
  ggplot(aes(year, value, color = type, 
             group = paste0(model, experiment, ensemble, type))) +
  geom_line() +
  facet_wrap(~experiment, scales = "free") +
  labs(x = "Year",
       y = "tas",
       title = "CMIP6 runs - tas over time (models with Tgav values)") +
  theme_minimal()

# Plot all models, even those without Tgav
tas$type = "land"
tas_l_plot <- tas %>%
  select(c(year, value, model, experiment, ensemble, type))

# Combine data
plot_data <- bind_rows(tas_l_plot, Tgav_plot)

# Plot
plot_data %>%
  # filter(model != "MPI-ESM1-2-LR") %>%
  ggplot(aes(year, value, color = type, 
             group = paste0(model, experiment, ensemble, type))) +
  geom_line() +
  facet_wrap(~experiment, scales = "free") +
  labs(x = "Year",
       y = "tas",
       title = "CMIP6 runs - tas over time") +
  theme_minimal()

# Plot just tas_land models (no global comparison)
tas %>%
  ggplot(aes(year, value, color = model, 
             group = paste0(model, experiment, ensemble))) +
  geom_line() +
  facet_wrap(~experiment, scales = "free") +
  labs(x = "Year",
       y = "tas",
       title = "CMIP6 runs - tas over time") +
  theme_minimal()

# What are the low models?
low_models <- tas %>% 
  filter(value < 200)
low_models <- unique(low_models$model)
low_models

